---
title: "ECG-Rough-Draft"
author:
  - name: Megan Osakwe
  - name: Shivansh Patel
  - name: Christopher Rios
  - name: Humberto Rodriguez
subtitle: |
  MATH4322
  SP26
format: 
  pdf:
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}

documentclass: report
papersize: letter
toc: true
geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm

---
\newpage
## Introduction

## Methods

### Decision Tree (Shivansh Patel & Christopher Rios)

**Base Model Formula** 
$$
\begin{split}
ECG\_signal &= hbpermin + Pseg + PQseg + QRSseg + QRseg + QTseg + RSseg \\
&+ STseg + Tseg + PTseg + ECGseg + RRmean + PPmean + PQdis \\
&+ PonQdis + PRdis + PonRdis + PSdis + PonSdis + PTdis + PonTdis \\
&+ PToffdis + QRdis + QSdis + QTdis + QToffdis + RSdis + RTdis \\
&+ RToffdis + STdis + SToffdis + PonToffdis + PQslope + STslope \\
&+ QRSarea + QRSperi + RRTot + NNTot + SDRR + IBIM + IBISD \\
&+ SDSD + RMSSD + NN50 + pNN50 
\end{split}
$$ {#eq-treeBase}

**Gini Index** 
$$
G = \sum_{k=1}^{K} \hat{p_{mk}}(1 - \hat{p_{mk}})
$$ {#eq-gini}

**Entropy** 
$$
D = -\sum_{k=1}^{K} \hat{p_{mk}}log\hat{p_{mk}}
$$ {#eq-entropy}

### Random Forest (Megan Osakwe & Humberto Rodriguez)

## Results

## Conclusion

## Bibliography

Alekhya Lanka. (2023). ECG of Cardiac Ailments Dataset [Data set]. Kaggle.  
https://doi.org/10.34740/KAGGLE/DSV/5917467

## Appendix

### Description of Variables Table

```{r}
#| label: tbl-variables
#| tbl-cap: "Description of ECG Variables"
#| warning: false
#| message: false
#| echo: false
library(knitr)
library(kableExtra)

raw_data <- read.csv("VariableDescriptions.csv")
variables.df <- raw_data[, 1:2] 
colnames(variables.df) <- c("Variable", "Description")

n_rows <- nrow(variables.df)
mid_point <- ceiling(n_rows / 2)

left_half <- variables.df[1:mid_point, ]
right_half <- variables.df[(mid_point + 1):n_rows, ]

if (nrow(left_half) > nrow(right_half)) {
    right_half <- rbind(right_half, data.frame(Variable = "", Description = ""))
}

split_table <- cbind(left_half, right_half)

kable(split_table, "latex", booktabs = FALSE, escape = TRUE, col.names = c("Variable", "Description", "Variable", "Description"), align = "p{2.0cm}p{6cm}|p{2.0cm}p{6cm}") %>%
    kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) %>%
    column_spec(2, border_right = TRUE)
```
\newpage

### Data Cleaning Source Code

```{r}
#| eval: false
#| results: false
# Import dataset, assuming file is in current working directory
ecg.df <- read.csv("ECGCvdata.csv")

# Response Variable counts
ailement_counts <- table(ecg.df$ECG_signal)
print(ailement_counts)

# Finding variables with empty values
cols_with_na <- names(ecg.df)[colSums(is.na(ecg.df)) > 0]
print(cols_with_na)

# Dropping columns
to_drop <- c("RECORD", "QRtoQSdur", "RStoQSdur", "PonPQang", "PQRang", "QRSang", "RSTang", "STToffang", "QRslope", "RSslope")
ecg.df <- ecg.df[, !(names(ecg.df) %in% to_drop)]

# Converting response to factor
ecg.df$ECG_signal <- as.factor(ecg.df$ECG_signal)
```

### Decision Tree Source Code

```{r}
#| eval: false
#| results: false
# Insert Source Code Here
1 + 2
```

### Random Forest Source Code

```{r}
#| eval: false
#| results: false
# Load neccesary packages
library(randomForest)
library(caret)

# Creates a vector to store the mean test error rates
mean_test_error.rf <- c();
for(i in c(1:10))
{

set.seed(i)

# Creating stratified data for training and test
train_index.rf <- createDataPartition(ecg.df$ECG_signal, p = 0.8, list = FALSE)
train_data.rf <- ecg.df[train_index.rf, ]
test_data.rf <- ecg.df[-train_index.rf, ]

#Fit the model
ecg.rf <- randomForest(ECG_signal ~ ., data = train_data.rf, ntree = 500, importance = TRUE)

#Predictions on test data and storing Mean Test Error
predictions.rf <- predict(ecg.rf, newdata = test_data.rf)
mean_test_error.rf[i] <- mean(predictions.rf != test_data.rf$ECG_signal)

}
```