---
title: "ECG-Rough-Draft"
author:
  - name: Megan Osakwe
  - name: Shivansh Patel
  - name: Christopher Rios
  - name: Humberto Rodriguez
subtitle: |
  MATH4322
  SP26
format: 
  pdf:
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}

documentclass: report
papersize: letter
toc: true
geometry:
      - top=20mm
      - left=20mm
      - right=20mm
      - bottom=20mm

---
\newpage
## Introduction

The electrocardiogram (ECG) was invented in 1902 by Dutch doctor and physiologist Williem Einthoven.The ECG records the heart's electrical signals using multiple electrodes placed on the surface of a persons skin. These electrodes are strategically positioned to monitor electrical activity in specific regions of the heart.The ECG is crucial in the initial evaluation of a 

## Methods

### Decision Tree (Shivansh Patel & Christopher Rios)

**Base Model Formula** 
$$
\begin{split}
ECG\_signal &= hbpermin + Pseg + PQseg + QRSseg + QRseg + QTseg + RSseg \\
&+ STseg + Tseg + PTseg + ECGseg + RRmean + PPmean + PQdis \\
&+ PonQdis + PRdis + PonRdis + PSdis + PonSdis + PTdis + PonTdis \\
&+ PToffdis + QRdis + QSdis + QTdis + QToffdis + RSdis + RTdis \\
&+ RToffdis + STdis + SToffdis + PonToffdis + PQslope + STslope \\
&+ QRSarea + QRSperi + RRTot + NNTot + SDRR + IBIM + IBISD \\
&+ SDSD + RMSSD + NN50 + pNN50 
\end{split}
$$ {#eq-treeBase}

**Gini Index** 
$$
G = \sum_{k=1}^{K} \hat{p_{mk}}(1 - \hat{p_{mk}})
$$ {#eq-gini}

**Entropy** 
$$
D = -\sum_{k=1}^{K} \hat{p_{mk}}log\hat{p_{mk}}
$$ {#eq-entropy}

### Random Forest (Megan Osakwe & Humberto Rodriguez)

[Rough Description of the methodology and what the code is supposed to accomplish - clean up]

While the decision tree methodology is useful for visualization and can handle highly dimension data, it suffers from high variance, since different subsets from the same dataset can lead to different results. The random forest methodology can similarly handle highly dimensional data, while also reducing variance by averaging the result of many decorrelated trees, leading to a model that is resistant to overfitting.  
[Tie in results of decision model here to see what random forest can improve on]  
The model formula for the random forest is the same as the decision tree, which is as follows:

$$
\begin{split}
ECG\_signal &= hbpermin + Pseg + PQseg + QRSseg + QRseg + QTseg + RSseg \\
&+ STseg + Tseg + PTseg + ECGseg + RRmean + PPmean + PQdis \\
&+ PonQdis + PRdis + PonRdis + PSdis + PonSdis + PTdis + PonTdis \\
&+ PToffdis + QRdis + QSdis + QTdis + QToffdis + RSdis + RTdis \\
&+ RToffdis + STdis + SToffdis + PonToffdis + PQslope + STslope \\
&+ QRSarea + QRSperi + RRTot + NNTot + SDRR + IBIM + IBISD \\
&+ SDSD + RMSSD + NN50 + pNN50 
\end{split}
$$ {#eq-treeBase}

```{r}
#| label: setup-rf
#| ref.label: !expr c("data-cleaning", "rf-packages", "rf-fit", "rf-importance-calc")
#| eval: true
#| echo: false
#| results: hide
#| message: false
#| warning: false
```

```{r}
#| ref.label: rf-importance-MDE-plot
#| eval: true
#| echo: false
#| results: markup
```


Model ran 10 times in a loop. Variables created to store importance and MTE of each run. Importance is a list because importance() produces a matrix, and vector c() couldnt accomodate. After model is fitted, calculations for both importance and mte is done using the variables created. if summary of ecg.rf was done, would only see the result of the last run, since each iteration overides the variable ecg.rf. varImpPlot() is not used for similar reasons, as it only provides a plot of the importance values of the last run. dotchart() can be used to achieve a plot with a similar layout as varImpPlot(). 

## Results

## Conclusion

## Bibliography

Alekhya Lanka. (2023). ECG of Cardiac Ailments Dataset [Data set]. Kaggle.  
https://doi.org/10.34740/KAGGLE/DSV/5917467

Sattar Y, Chhabra L. Electrocardiogram. [Updated 2023 Jun 5]. In: StatPearls [Internet]. Treasure Island (FL): StatPearls Publishing; 2026 Jan-.  
https://www.ncbi.nlm.nih.gov/books/NBK549803/

## Appendix

### Description of Variables Table

```{r}
#| label: tbl-variables
#| tbl-cap: "Description of ECG Variables"
#| warning: false
#| message: false
#| echo: false
library(knitr)
library(kableExtra)

raw_data <- read.csv("VariableDescriptions.csv")
variables.df <- raw_data[, 1:2] 
colnames(variables.df) <- c("Variable", "Description")

n_rows <- nrow(variables.df)
mid_point <- ceiling(n_rows / 2)

left_half <- variables.df[1:mid_point, ]
right_half <- variables.df[(mid_point + 1):n_rows, ]

if (nrow(left_half) > nrow(right_half)) {
    right_half <- rbind(right_half, data.frame(Variable = "", Description = ""))
}

split_table <- cbind(left_half, right_half)

kable(split_table, "latex", booktabs = FALSE, escape = TRUE, col.names = c("Variable", "Description", "Variable", "Description"), align = "p{2.0cm}p{6cm}|p{2.0cm}p{6cm}") %>%
    kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) %>%
    column_spec(2, border_right = TRUE)
```
\newpage

### Data Cleaning Source Code

```{r}
#| label: data-cleaning
#| eval: false
#| results: false
#| message: false
#| warning: false
# Import dataset, assuming file is in current working directory
ecg.df <- read.csv("ECGCvdata.csv")

# Response Variable counts
ailement_counts <- table(ecg.df$ECG_signal)
print(ailement_counts)

# Finding variables with empty values
cols_with_na <- names(ecg.df)[colSums(is.na(ecg.df)) > 0]
print(cols_with_na)

# Dropping columns
to_drop <- c("RECORD", "QRtoQSdur", "RStoQSdur", "PonPQang", "PQRang", "QRSang", "RSTang", "STToffang", "QRslope", "RSslope")
ecg.df <- ecg.df[, !(names(ecg.df) %in% to_drop)]

# Converting response to factor
ecg.df$ECG_signal <- as.factor(ecg.df$ECG_signal)
```

### Decision Tree Source Code

```{r}
#| eval: false
#| results: false
# Insert Source Code Here
1 + 2
```

### Random Forest Source Code

```{r}
#| label: rf-packages
#| eval: false
#| results: false
# Load necessary packages
library(randomForest)
library(caret)
```
```{r}
#| label: rf-fit
#| eval: false
#| results: false
# Creating variables to store variable importance and MTE
p <- ncol(ecg.df) - 1
importance.rf <- list()
mte.rf <- c();

for(i in c(1:10))
{

set.seed(i)

# Creating stratified data for training and test
train_index.rf <- createDataPartition(ecg.df$ECG_signal, p = 0.8, list = FALSE)
train_data.rf <- ecg.df[train_index.rf, ]
test_data.rf <- ecg.df[-train_index.rf, ]

# Fit the model
ecg.rf <- randomForest(ECG_signal ~ ., data = train_data.rf, mtry = sqrt(p), ntree = 500, importance = TRUE)

# Predictions on test data and storing Mean Test Error
predictions.rf <- predict(ecg.rf, newdata = test_data.rf)
mte.rf[i] <- mean(predictions.rf != test_data.rf$ECG_signal)

# Creating data frame for importance
importance.df <- as.data.frame(importance(ecg.rf))

# Creating a column for the variable to stop splitting of results per run
importance.df$Feature <- rownames(importance.df)

# Storing variable importance
importance.rf[[i]] <- importance.df
}
```
```{r}
#| label: rf-importance-calc
#| eval: false
#| results: false
# Importance Calculations 

# Aggregate the importance score
all.importance.rf <- do.call(rbind, importance.rf)
avg.importance.rf <- aggregate(. ~ Feature, data = all.importance.rf, FUN = mean)

# Sorting the results for list of top variables
avg.importance.rf <- avg.importance.rf[order(-avg.importance.rf$MeanDecreaseAccuracy), ]
head(avg.importance.rf, 10)
```
```{r}
#| label: rf-accuracy-calc
#| eval: false
#| results: false
# Accuracy Calculations

# Averaging the accuracy across all runs
avg.accuracy.rf <- 1 - mean(mte.rf)

# Standard Deviation of accuracy across all runs
sd.accuracy.rf <- sd(1 - mte.rf)

# Confusion matrix for comparison
confusion.rf <- confusionMatrix(predictions.rf, test_data.rf$ECG_signal)
```
```{r}
#| label: rf-results
#| eval: false
#| results: false
# Print results
print(paste("Average Accuracy across 10-fold split:", round(avg.accuracy.rf, 4)))
print(paste("Standard Deviation of accuracy across 10-fold split:", round(sd.accuracy.rf, 4)))
print(confusion.rf)
```
```{r}
#| label: rf-importance-MDE-plot
#| eval: false
#| results: false
# Sorting the results to plot the average Mean Decrease Accuracy 
avg.importance.rf <- avg.importance.rf[order(avg.importance.rf$MeanDecreaseAccuracy), ]
top20.rf <- tail(avg.importance.rf, 20)
dotchart(top20.rf$MeanDecreaseAccuracy, 
         labels = top20.rf$Feature,
         cex = 0.7, 
         main = "Average Variable Importance",
         xlab = "Mean Decrease Accuracy")
```